---
title: "Par Lab prior fit: per-shape par deviations"
author: "Christopher D. Wells"
date: 2026-08-03
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
---

This document fits the completed Par Lab battery (86 boards, played 2026-08-02
and 2026-08-03) and produces posterior centers, with honest widths, for the
`PAR_MODEL_SHAPES` deviation terms in `src/logic/difficulty.js`. It also
produces the par-per-config tables the Challenge-250 tier ladder and the
Phase-2 tiling par-band configs need. Nothing here writes into the shipped
model. The final section lays out the options for how these priors could enter
the nightly refit; that choice is a design decision made outside this document.

The deviation vocabulary is the nightly refit's own: for each tiling, an
intercept deviation (`shapeHex`, `shapeCairo`, ...) plus per-feature slope
deviations (`shapeHex_x_cellCount`, ...), all signed log-scale shifts from the
square equation. A deviation of 0 means the shape prices exactly as a square
board with the same feature vector; the shipped blocks today are all at 0.
The lab exists because live daily data cannot identify the size and density
deviations at all (the rotation ships one fixed config per shape, so those
columns are exactly collinear with the shape's intercept deviation), and
identifies everything else slowly.

Rendering takes a few minutes: five small Stan models compile and sample.

```{r setup}
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(knitr)
library(brms)
library(posterior)

options(brms.backend = "cmdstanr")
set.seed(20260803)

`%||%` <- function(a, b) if (is.null(a)) b else a
geo_mean <- function(x) exp(mean(log(x)))
```

## 1. Constants and the frozen data snapshot

Two tables are mirrored verbatim from `scripts/refit-par-model.R` so this
document speaks the nightly refit's names: `COEF_TO_PREDICTOR` (JS coefficient
key to model predictor) and `SHAPE_TABLE` (tiling type string to deviation-term
stem). The lockstep between those tables and the client registry is pinned by
`test/tilingParModelContract.test.mjs`; if a seventh tiling ever lands, both
copies grow together.

The row data is a frozen snapshot of the world-readable `parLab/` node, pulled
2026-08-03 with the battery complete at 86/86. The chunk re-downloads only if
the snapshot file is missing, so a re-render is byte-stable and needs no
network.

```{r constants}
DIFFICULTY_PATH <- "../src/logic/difficulty.js"
SNAPSHOT_PATH   <- "data/parlab-rows-2026-08-03.json"
PARLAB_URL      <- "https://gregsweeper-66d02-default-rtdb.firebaseio.com/parLab.json"
HIS_UID_PREFIX  <- "V07QPXYa"

COEF_TO_PREDICTOR <- c(
  secPerCell         = "cellCount",
  secPerMineFlag     = "totalMines",
  secPerPatternMove  = "patternMoves",
  secPerSearchMove   = "searchMoves",
  secPerWallEdge     = "wallEdgeCount",
  secPerZeroCluster  = "zeroClusterCount",
  secPerMysteryCell  = "mysteryCellCount",
  secPerLiarCell     = "liarCellCount",
  secPerLockedCell   = "lockedCellCount",
  secPerWormholePair = "wormholePairCount",
  secPerMirrorPair   = "mirrorPairCount",
  secPerSonarCell    = "sonarCellCount",
  secPerCompassCell  = "compassCellCount",
  secPerWormLoad     = "wormLoad"
)

SHAPE_TABLE <- c(
  "4.8.8"     = "shape488",
  "hex"       = "shapeHex",
  "cairo"     = "shapeCairo",
  "floret"    = "shapeFloret",
  "rhombille" = "shapeRhombille",
  "deltoidal" = "shapeDeltoidal"
)

# The gimmick features the modifier singles placed. Mystery, mirror, and walls
# were cut from the battery at the board-66 trim, so their columns are all-zero
# here and never enter the fit.
GIMMICK_FEATURES <- c("sonarCellCount", "compassCellCount", "wormholePairCount",
                      "liarCellCount", "lockedCellCount", "wormLoad")

if (!file.exists(SNAPSHOT_PATH)) {
  dir.create(dirname(SNAPSHOT_PATH), showWarnings = FALSE)
  download.file(PARLAB_URL, SNAPSHOT_PATH, mode = "wb", quiet = TRUE)
}
raw <- fromJSON(SNAPSHOT_PATH, simplifyVector = FALSE)
cat("snapshot rows:", length(raw), "\n")
```

## 2. The base equation the deviations are measured against

The shipped `PAR_MODEL` block is parsed from `difficulty.js` at render time
(the same marker-parse the nightly uses), never hand-copied, so there is no
transcription to drift. Every deviation in this document is defined relative
to the coefficients printed below. If a later refit moves the base and this
document is re-rendered, the deviations shift with it; the vintage line makes
that visible.

```{r base-model}
src <- paste(readLines(DIFFICULTY_PATH, warn = FALSE, encoding = "UTF-8"),
             collapse = "\n")
bs <- str_locate(src, fixed("// PAR_MODEL:START"))[1, "end"]
be <- str_locate(src, fixed("// PAR_MODEL:END"))[1, "start"]
stopifnot(!is.na(bs), !is.na(be))
block <- substr(src, bs + 1, be - 1)
stopifnot(str_detect(block, "scale\\s*:\\s*'log'"))
kv <- str_match_all(block, "(\\w+)\\s*:\\s*(-?[0-9]+(?:\\.[0-9]+)?)")[[1]]
PAR_MODEL <- setNames(as.list(as.numeric(kv[, 3])), kv[, 2])
base_vintage <- str_match(block, "Last refit: (\\d{4}-\\d{2}-\\d{2})")[1, 2]

stopifnot(all(names(COEF_TO_PREDICTOR) %in% names(PAR_MODEL)))
cat("base model vintage (last refit):", base_vintage, "\n")
kable(tibble(coefficient = c("intercept", names(COEF_TO_PREDICTOR)),
             value = as.numeric(PAR_MODEL[c("intercept", names(COEF_TO_PREDICTOR))])),
      digits = 5)
```

## 3. Rows, and the agreed conventions

Every convention below was agreed before the battery was played, and each has
a client-side test pinning it. In order:

1. **Filter to one player.** Every row is Christopher's; the uid filter is a
   guard, not a selection.
2. **Tombstones first.** A `result: 'invalid'` row voids every win or skip
   with the same `(id, attempt)`. One exists: the board-15 warm-up
   (`w-floret-3`, attempt 0), a deliberately mine-popped board.
3. **Probe filter.** Any row detonating more than 30% of the board's mines is
   a layout probe, not a solve (the daily fit's rule). On this data the only
   row past the bar is the same tombstoned board, so the two guards agree.
4. **Pure time.** `timeSec` on rows carrying `penaltySec` is the daily
   convention, penalty-inclusive; pure play time is `timeSec - penaltySec`.
   Rows without the field are pure wall clock already, for either of two
   reasons: recorded before the 2026-08-02 fix (integer wall clock, penalties
   never folded in), or recorded after it with zero strikes (`penaltySec` is
   only written when positive). Both read verbatim, so the rule collapses to
   one line: subtract `penaltySec`, treating a missing field as 0.
5. **Warm-ups are excluded from estimates** but their per-shape learning
   slopes are checked in section 5, and warm-up plays still count toward each
   shape's exposure when the practice covariate is built.

A note on why pure time removes the whole recorded penalty rather than the
nightly's base-only recipe. The daily fit keeps each strike's info-value in
the response, because that part of the penalty stands in for deduction work
the player skipped, priced by the model. Here that pricing would be circular:
the lab's strike info-values were computed under the square model, which is
exactly the model this fit is measuring the lattices against. So the response
is honest wall clock with the whole penalty removed, and a strike's skipped
deduction work is left as a small fast-bias on strike-heavy rows. Section 9's
sensitivity S4 measures that bias directly, and re-pricing the info-values
under the fitted per-shape model (fit, re-price, refit) remains available as a
second pass if S4 shows it matters.

```{r rows}
rows <- tibble(pushId = names(raw), entry = unname(raw)) |>
  mutate(
    uid        = map_chr(entry, ~ .x$uid %||% NA_character_),
    id         = map_chr(entry, ~ .x$id %||% NA_character_),
    attempt    = map_int(entry, ~ as.integer(.x$attempt %||% 0)),
    shape      = map_chr(entry, ~ .x$shape %||% NA_character_),
    result     = map_chr(entry, ~ .x$result %||% NA_character_),
    warmup     = map_lgl(entry, ~ isTRUE(.x$warmup)),
    gimmick    = map_chr(entry, ~ paste(unlist(.x$gimmicks %||% list()), collapse = "+")),
    timeSec    = map_dbl(entry, ~ .x$timeSec %||% NA_real_),
    penaltySec = map_dbl(entry, ~ .x$penaltySec %||% 0),
    bombHits   = map_dbl(entry, ~ .x$bombHits %||% 0),
    seq        = map_int(entry, ~ as.integer(.x$seq %||% NA_integer_)),
    par_row    = map_dbl(entry, ~ .x$par %||% NA_real_),
    playedAt   = map_chr(entry, ~ .x$playedAt %||% NA_character_),
    day        = substr(playedAt, 1, 10),
    wormRealized = map_dbl(entry, ~ {
      ev <- .x$wormEvents %||% list()
      if (length(ev) == 0) return(NA_real_)
      sum(map_dbl(ev, ~ (.x$len %||% 0) * (.x$moves %||% 0))) / 100
    }),
    features   = map(entry, ~ .x$features)
  ) |>
  filter(startsWith(uid, HIS_UID_PREFIX))

# Battery-shape invariants: 86 boards, the four blocks at their post-trim,
# post-addendum sizes, no unhandled result kinds.
stopifnot(
  nrow(rows) > 0,
  n_distinct(rows$id) == 86,
  all(rows$result %in% c("win", "invalid")),
  sum(rows$result == "invalid") == 1
)
prefix_counts <- rows |> distinct(id) |>
  count(prefix = str_extract(id, "^[a-z]+")) |>
  arrange(prefix)
stopifnot(identical(
  prefix_counts,
  tibble(prefix = c("a", "m", "p", "w"), n = c(7L, 14L, 47L, 18L))
))

# Tombstones void their (id, attempt); then wins only.
voided_keys <- rows |> filter(result == "invalid") |> mutate(k = paste(id, attempt)) |> pull(k)
all_plays <- rows |> filter(result == "win")   # includes the voided probe: it was still exposure
wins <- all_plays |> filter(!paste(id, attempt) %in% voided_keys)

# Features ride every win row; unnest them wide.
feats <- wins |> select(pushId, features) |> unnest_wider(features)
d <- wins |> select(-features) |> inner_join(feats, by = "pushId") |>
  mutate(across(c(cellCount, totalMines, density, passAMoves, canonicalSubsetMoves,
                  genericSubsetMoves, advancedLogicMoves, wallEdgeCount,
                  zeroClusterCount, mysteryCellCount, liarCellCount, lockedCellCount,
                  wormholePairCount, mirrorPairCount, sonarCellCount,
                  compassCellCount, wormLoad),
                ~ ifelse(is.na(.x), 0, as.numeric(.x))))
stopifnot(nrow(d) == 86, !anyNA(d$timeSec), !anyNA(d$seq))

# The probe filter (mirrors isBombHitCheat): on this battery it catches only
# the already-tombstoned board, so nothing further drops.
d <- d |> mutate(probe = bombHits > 0.30 * totalMines)
stopifnot(sum(d$probe) == 0)

# Pure time, and the worm realized dose. wormLoad becomes the REALIZED
# segment-moves (sum of len x moves / 100 from the row's own wormEvents),
# matching the nightly's realized-dose convention; the scheduled maximum is
# kept aside for the realization table. Only the four worm singles carry
# events, and a fast solve or a late hatch realizes far less than the
# schedule, which is the whole case for the instrument.
d <- d |>
  mutate(
    pure_time     = pmax(timeSec - penaltySec, 1),
    wormScheduled = wormLoad,
    wormLoad      = ifelse(!is.na(wormRealized), wormRealized, 0)
  )
stopifnot(all(d$pure_time > 0), sum(!is.na(d$wormRealized)) == 4)

# Derived reasoning tiers, exactly as the nightly derives them.
d <- d |> mutate(
  patternMoves = canonicalSubsetMoves + genericSubsetMoves,
  searchMoves  = advancedLogicMoves
)

cat(nrow(d), "usable win rows:",
    sum(d$warmup), "warm-ups (learning check only) +",
    sum(!d$warmup), "measured\n")
```

## 4. Parity par: the offset every row is compared against

Each row's features are priced under the base model printed in section 2,
giving the parity par: what the shipped `predictPar` says this exact board is
worth when the shape deviations are all zero. The response the model fits is
`log(pure_time / par_sq)`, so a coefficient of 0 anywhere means "prices like a
square board."

The rows also carry the par the client computed at play time. Those two
disagree slightly on rows played before the 2026-08-02 nightly refit landed
(a different coefficient vintage); recomputing every row under one vintage is
the point, and the table below shows the drift is a few seconds at most.

```{r parity-par}
lp <- rep(PAR_MODEL$intercept, nrow(d))
for (k in names(COEF_TO_PREDICTOR)) {
  lp <- lp + PAR_MODEL[[k]] * d[[COEF_TO_PREDICTOR[[k]]]]
}
d$par_sq    <- exp(lp)
d$ratio     <- d$pure_time / d$par_sq
d$log_ratio <- log(d$ratio)

d |>
  mutate(drift = abs(par_row - par_sq)) |>
  summarise(
    rows_matching_exactly = sum(drift < 0.05, na.rm = TRUE),
    rows_drifted          = sum(drift >= 0.05, na.rm = TRUE),
    max_drift_sec         = max(drift, na.rm = TRUE)
  ) |>
  kable(digits = 1)
```

## 5. Warm-up learning check

Three consecutive plain boards per shape at its daily config, played easiest
lattice first, before any measured row. The question the design left open:
did each shape's curve flatten by board three? These rows never enter the
estimates either way.

```{r warmups}
warm <- d |> filter(warmup) |> arrange(seq) |>
  group_by(shape) |> mutate(round = row_number()) |> ungroup()

warm |>
  select(shape, round, seq, pure_time, par_sq, ratio, bombHits) |>
  kable(digits = 2)

ggplot(warm, aes(round, ratio, color = shape)) +
  geom_line() + geom_point() +
  scale_y_log10() +
  labs(x = "warm-up round", y = "pure time / parity par (log scale)",
       title = "Warm-up trajectories by shape") +
  theme_minimal()
```

The read: hex fell to about half par and stayed there; 4.8.8 held near his
square pace throughout; cairo and deltoidal came back down after a slow middle
board; rhombille had the steepest drop and kept falling. Floret is the
outlier in the other direction: its third warm-up (the fresh layout issued
after the voided probe) was its slowest, so floret entered the measured
battery still learning. That is the direct motivation for the practice term
in section 7, and the reason floret's deviations get the least-settled
language in the results.

The replicate pairs make the same point inside the measured rows: the same
size and density cell, played again 30 to 50 boards later, came back far
faster on the two lattices that were still noisy.

```{r replicates}
meas <- d |> filter(!warmup)

reps <- meas |> filter(str_ends(id, "r")) |> mutate(base_id = str_remove(id, "r$"))
reps |>
  inner_join(meas |> select(base_id = id, seq_first = seq, ratio_first = ratio),
             by = "base_id") |>
  select(config = base_id, seq_first, ratio_first, seq_again = seq, ratio_again = ratio) |>
  kable(digits = 2)
```

A halving on deltoidal and a factor of 6 to 10 on floret is some mix of
practice and layout luck; at one replicate per cell the two cannot be
separated, and the model's practice term plus its honest residual widths are
how that uncertainty is carried rather than hidden.

## 6. First look: anchors, grid, and raw multipliers

The seven square anchors are the bridge to the base model: they measure what
fraction of Greg par Christopher's lab pace runs, so every shape read is
relative to his own square pace rather than to par directly. His shipped
handicap (k = 0.971 across all his daily play) does not transfer to a focused
lab sitting, and the anchors are the within-session instrument that replaces
it.

```{r anchors}
anchors <- meas |> filter(shape == "rect")
anchors |>
  select(id, seq, day, timeSec, penaltySec, pure_time, par_sq, ratio) |>
  kable(digits = 2)
cat("anchor geometric mean ratio:", round(geo_mean(anchors$ratio), 3), "\n")
```

Anchor a-01 (0.32, the battery's first measured board) is well off the other
six; the day-two anchors sit between 0.51 and 0.88 with no trend across 50
boards, so a single session-pace intercept is a reasonable primary and
sensitivity S2 checks a drift term. Raw geometric-mean ratios by shape, with
the anchor pace as the yardstick:

```{r raw-read}
raw_read <- meas |>
  group_by(shape) |>
  summarise(n = n(), geo_ratio = geo_mean(ratio), log_sd = sd(log_ratio),
            .groups = "drop") |>
  mutate(vs_anchor_pace = geo_ratio / geo_mean(anchors$ratio)) |>
  arrange(vs_anchor_pace)
kable(raw_read, digits = 2)

ggplot(meas |> filter(shape != "rect", str_starts(id, "p-")),
       aes(cellCount, ratio, color = factor(round(density, 2)))) +
  geom_point(size = 2.5) +
  scale_y_log10() +
  facet_wrap(~shape) +
  labs(x = "cells", y = "pure time / parity par (log scale)", color = "density",
       title = "Plain grid: where each lattice departs from square pricing") +
  theme_minimal()
```

This is the completion read the fit has to formalize: hex prices at about his
square pace, 4.8.8 a third dearer, rhombille and cairo about double, floret
wide and learning-contaminated, deltoidal near five times square par with its
worst cells at high density rather than large size.

## 7. The model

One Bayesian regression on the 68 measured rows (47 plain grid, 14 modifier
singles, 7 anchors), response `log(pure_time / par_sq)`:

- **Intercept**: his lab session pace on square boards, identified by the
  anchors (all deviation columns are zero on rect rows).
- **practice_c**: log of each tiling row's within-shape exposure (warm-ups
  and the voided probe count as exposure), centered on the measured tiling
  rows' mean, zero on rect rows. One shared slope: the power law of practice,
  with squares exempt because they are his native game. Centering makes every
  shape deviation read at battery-average proficiency rather than at either
  extreme; section 10 returns to what that means for a live novice.
- **Per-shape core deviations**, three per lattice in the nightly's exact
  column names: the intercept deviation (`shapeHex`, ...), the per-cell slope
  deviation (`shapeHex_x_cellCount`, ...), and the per-mine slope deviation
  (`shapeHex_x_totalMines`, ...). These are what the size-by-density grid was
  designed to identify, and they are the deliverable.
- **Gimmick deviations** where the mechanism block placed a board, again in
  the nightly's names (`shapeHex_x_sonarCellCount`, `shapeRhombille_x_wormLoad`,
  ...): fourteen columns, each nonzero on exactly one row. These are
  observations wearing coefficient clothes, and the results section treats
  them that way.

What is deliberately absent: per-shape deviations on the reasoning tiers
(`patternMoves`, `searchMoves`) and on `zeroClusterCount`. Those features vary
across the lab boards, but within a shape they move with size and density, and
six to eleven grid rows per shape cannot separate five slopes. The parity
offset already charges them at square rates; whatever a lattice does to their
true cost lands in the three fitted core terms for now, and the residual
diagnostic in section 8 checks whether that compromise leaves structure on
the table. Their nightly deviations stay zero-centered until live data can
speak to them.

Priors, all on the log scale:

- Core deviations: `normal(0, 0.5)`, the nightly's own `INTERACTION_PRIOR_SD`.
  Zero-centered is the "a shape's equation starts as the square's" design; 0.5
  is wide here (a shape would need to quadruple par before 2 SD), and for the
  per-unit slopes it is effectively flat.
- Gimmick deviations: `normal(0, 1)` per feature unit, wide by instruction:
  with one board per cell the posterior is the observation, barely shrunk.
- Intercept `normal(0, 1)`, practice slope `normal(0, 0.5)`, residual
  `normal(0, 1)` half-normal: weakly informative throughout.

All terms are signed, so there is no positivity bound anywhere and none of the
boundary geometry that bit the nightly's 2026-08-01 run; this is plain smooth
sampling.

```{r design}
# Within-shape exposure order (warm-ups and the voided probe both count: each
# was a played board and practice accrued), then the centered practice axis.
exposure <- all_plays |>
  filter(shape != "rect") |>
  arrange(seq) |>
  group_by(shape) |>
  mutate(order = row_number()) |>
  ungroup() |>
  select(pushId, order)

fit_df <- meas |>
  left_join(exposure, by = "pushId") |>
  mutate(practice_raw = ifelse(shape == "rect", NA_real_, log(order)))
practice_center <- mean(fit_df$practice_raw, na.rm = TRUE)
fit_df <- fit_df |>
  mutate(practice_c = ifelse(is.na(practice_raw), 0, practice_raw - practice_center))

# Deviation columns, exact nightly names: indicator, indicator x cells,
# indicator x mines, indicator x each gimmick feature.
for (js in names(SHAPE_TABLE)) {
  stem <- SHAPE_TABLE[[js]]
  ind  <- as.numeric(fit_df$shape == js)
  fit_df[[stem]] <- ind
  fit_df[[paste0(stem, "_x_cellCount")]]  <- ind * fit_df$cellCount
  fit_df[[paste0(stem, "_x_totalMines")]] <- ind * fit_df$totalMines
  for (g in GIMMICK_FEATURES) {
    fit_df[[paste0(stem, "_x_", g)]] <- ind * fit_df[[g]]
  }
}
core_cols <- c(unname(SHAPE_TABLE),
               paste0(rep(unname(SHAPE_TABLE), each = 2), "_x_",
                      c("cellCount", "totalMines")))
gim_cols_all <- as.vector(outer(unname(SHAPE_TABLE), GIMMICK_FEATURES,
                                paste, sep = "_x_"))
gim_cols <- gim_cols_all[vapply(gim_cols_all,
                                function(cn) any(fit_df[[cn]] > 0), logical(1))]
stopifnot(length(core_cols) == 18, length(gim_cols) == 14)
cat("active gimmick columns:\n"); print(gim_cols)

model_rhs <- paste(c("0 + Intercept", "practice_c", core_cols, gim_cols),
                   collapse = " + ")
f_main <- as.formula(paste("log_ratio ~", model_rhs))

priors <- c(
  set_prior("normal(0, 1)",   class = "b", coef = "Intercept"),
  set_prior("normal(0, 0.5)", class = "b", coef = "practice_c"),
  do.call(c, lapply(core_cols, function(cn)
    set_prior("normal(0, 0.5)", class = "b", coef = cn))),
  do.call(c, lapply(gim_cols, function(cn)
    set_prior("normal(0, 1)", class = "b", coef = cn))),
  set_prior("normal(0, 1)", class = "sigma")
)
```

```{r fit-main, results="hide"}
fit_main <- brm(
  f_main, data = fit_df, prior = priors,
  chains = 4, iter = 3000, warmup = 1500,
  control = list(adapt_delta = 0.99),
  cores = 4, seed = 20260803, refresh = 0
)
```

## 8. Diagnostics

Same gates as the nightly: every Rhat at or under 1.05, every bulk ESS at or
over 400, divergent transitions under 0.25% of draws. The render stops here if
any gate fails, so a bad fit cannot quietly reach the results.

```{r diagnostics}
ps <- summarise_draws(as_draws_array(fit_main), "rhat", "ess_bulk")
n_div <- sum(nuts_params(fit_main)$Value[nuts_params(fit_main)$Parameter == "divergent__"])
total_draws <- 4 * 1500
cat(sprintf("max Rhat = %.3f, min bulk ESS = %.0f, divergent = %d/%d\n",
            max(ps$rhat, na.rm = TRUE), min(ps$ess_bulk, na.rm = TRUE),
            n_div, total_draws))
stopifnot(max(ps$rhat, na.rm = TRUE) <= 1.05,
          min(ps$ess_bulk, na.rm = TRUE) >= 400,
          n_div / total_draws <= 0.0025)

pp_check(fit_main, ndraws = 50) +
  labs(title = "Posterior predictive check on log(pure time / parity par)")
```

```{r residual-diagnostics}
fit_df$resid <- fit_df$log_ratio - colMeans(posterior_epred(fit_main))

ggplot(fit_df |> filter(shape != "rect"), aes(order, resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point(aes(color = shape)) +
  geom_smooth(method = "loess", se = FALSE, color = "grey40", span = 1) +
  labs(x = "within-shape exposure order", y = "residual (log scale)",
       title = "Residuals against exposure: is one practice slope enough?") +
  theme_minimal()

ggplot(fit_df |> filter(shape != "rect", str_starts(id, "p-")),
       aes(searchMoves, resid, color = shape)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point() +
  labs(x = "search moves (charged at square rates in the offset)",
       y = "residual (log scale)",
       title = "Residuals against search moves: the deviation the model omits") +
  theme_minimal()
```

If the second panel shows a clear per-shape ramp, the omitted reasoning-tier
deviations are real and the three core terms are absorbing them imperfectly;
that would argue for revisiting this fit once live data can price the tiers
per shape. A flat cloud says the compromise held at this sample size.

## 9. Results

### 9.1 The deviation posteriors

The deliverable table. Every term is a signed log-scale shift from the square
equation, in the exact name the nightly refit and `PAR_MODEL_SHAPES` use.
`n_rows` is how many fit rows carry the term; treat any n = 1 line as an
observation, not an estimate. The three core terms within a shape are jointly
identified and strongly correlated (an intercept shift can trade against the
two slopes across the grid's narrow config range), so their marginal widths
overstate the uncertainty of any board-level prediction; the config table in
9.2 is the well-identified combination.

```{r posterior-table}
fe <- fixef(fit_main)
dev_terms <- c(core_cols, gim_cols)
post_tbl <- tibble(
  term   = rownames(fe),
  mean   = fe[, "Estimate"],
  sd     = fe[, "Est.Error"],
  q2.5   = fe[, "Q2.5"],
  q97.5  = fe[, "Q97.5"]
) |>
  filter(term %in% c("Intercept", "practice_c", dev_terms)) |>
  mutate(n_rows = map_int(term, ~ if (.x %in% dev_terms)
    sum(fit_df[[.x]] != 0) else nrow(fit_df))) |>
  arrange(match(term, c("Intercept", "practice_c", dev_terms)))
kable(post_tbl, digits = 4)
```

```{r pace-note}
cat(sprintf("session pace: exp(Intercept) = %.2f of Greg par (the anchors' read)\n",
            exp(fe["Intercept", "Estimate"])))
cat(sprintf("practice slope: %.3f per log unit of exposure (negative = later boards faster)\n",
            fe["practice_c", "Estimate"]))
```

The deviations are pace-relative, so his session speed cancels out of every
shape read. What does not cancel is the standing assumption the ratio
handicap already makes: that his relative cost across lattices mirrors the
population's. One player's battery cannot test that; the live rotation will.

### 9.2 Composed multipliers per config

For each shape and config, the posterior of
`exp(shape + cells x shape_x_cellCount + mines x shape_x_totalMines)`: the
factor by which the lattice multiplies parity par at battery-average
proficiency. `par_sq` is the parity par of the actual generated board (for
the daily config, the mean over the three warm-up layouts), `fitted_par` is
its product with the multiplier, and `par_per_cell` is the Challenge-250
currency. Everything derives from the same posterior draws, so the intervals
are honest.

```{r config-table}
draws <- as_draws_df(fit_main)

mult_draws <- function(js, cells, mines) {
  stem <- SHAPE_TABLE[[js]]
  exp(draws[[paste0("b_", stem)]] +
      cells * draws[[paste0("b_", stem, "_x_cellCount")]] +
      mines * draws[[paste0("b_", stem, "_x_totalMines")]])
}

grid_cfgs <- meas |>
  filter(str_starts(id, "p-")) |>
  mutate(slot = str_remove(str_remove(id, "^p-[a-z0-9.]+-"), "r$")) |>
  group_by(shape, slot) |>
  summarise(cells = first(cellCount), mines = first(totalMines),
            density = first(density), par_sq = mean(par_sq),
            observed_geo_ratio = geo_mean(ratio), n_boards = n(),
            .groups = "drop")
daily_cfgs <- d |>
  filter(warmup) |>
  group_by(shape) |>
  summarise(slot = "daily", cells = first(cellCount), mines = first(totalMines),
            density = first(density), par_sq = mean(par_sq),
            observed_geo_ratio = NA_real_, n_boards = 0L, .groups = "drop")

# obs_mult is the model-free read of the same rows, made comparable to the
# fitted multiplier by dividing his fitted session pace out of the observed
# ratio; mult and obs_mult should track each other, with practice adjustment
# and shrinkage explaining the gaps.
config_tbl <- bind_rows(daily_cfgs, grid_cfgs) |>
  rowwise() |>
  mutate(
    mult      = mean(mult_draws(shape, cells, mines)),
    mult_lo   = quantile(mult_draws(shape, cells, mines), 0.025),
    mult_hi   = quantile(mult_draws(shape, cells, mines), 0.975),
    fitted_par = par_sq * mult,
    par_per_cell = fitted_par / cells,
    obs_mult  = observed_geo_ratio / exp(fe["Intercept", "Estimate"])
  ) |>
  ungroup() |>
  arrange(shape, cells, mines)

config_tbl |>
  select(shape, slot, cells, mines, density, par_sq, mult, mult_lo, mult_hi,
         fitted_par, par_per_cell, obs_mult) |>
  kable(digits = 2)
```

```{r daily-summary}
daily_mult <- config_tbl |> filter(slot == "daily") |>
  select(shape, mult, mult_lo, mult_hi, fitted_par)
kable(daily_mult |> arrange(mult), digits = 2)
```

Two sanity reads against the pre-fit numbers. First, the daily-config
multipliers should land near the raw completion read (section 6's
`vs_anchor_pace`, which is the same quantity computed without the model); a
large gap would mean the practice term or the slopes moved something, and the
sensitivity table in section 10 shows which. Second, deltoidal's density
story: its per-mine deviation against its per-cell deviation is the
model's answer to "density-dominated, not size-dominated," and the config
table's spread across its S2/M2/L2 column against its L0 row is the direct
check.

### 9.3 Modifier singles: observations, not conclusions

One board per gimmick-by-lattice cell, by design. Each line below is that one
board: the per-unit deviation posterior (wide, barely shrunk from the data),
the whole-board factor it implies at that board's feature count, and the raw
ratio against the shape's own plain fit. Nothing here is an estimate of a
population effect; these are priced observations that say which cells are
worth live boards first.

```{r gimmick-table}
gim_rows <- meas |> filter(str_starts(id, "m-"))
gim_tbl <- map_dfr(gim_cols, function(cn) {
  stem_f  <- str_split_1(cn, "_x_")
  js      <- names(SHAPE_TABLE)[SHAPE_TABLE == stem_f[1]]
  feature <- stem_f[2]
  row     <- gim_rows |> filter(shape == js, .data[[feature]] > 0)
  stopifnot(nrow(row) == 1)
  x <- row[[feature]]
  dr <- draws[[paste0("b_", cn)]]
  tibble(
    term = cn, board = row$id, x_units = x,
    dev_per_unit = mean(dr), dev_sd = sd(dr),
    board_factor = mean(exp(dr * x)),
    factor_lo = quantile(exp(dr * x), 0.025),
    factor_hi = quantile(exp(dr * x), 0.975),
    raw_ratio = row$ratio
  )
})
kable(gim_tbl |> arrange(term), digits = 3)
```

The worm four additionally carry their realized-dose story: the schedule is
what `predictPar` can know before play, the realization is what the run
experienced, and the gap between them (22% to 94% here) is why the fit prices
the realized dose. Rhombille's near-full dose with the battery's largest worm
slowdown is consistent with every covered number being load-bearing on the
lattice with no subset ladder; cairo's tiny dose with no slowdown is the same
mechanism from the other side. Two boards, so: consistent with, not shown.

```{r worm-table}
meas |>
  filter(wormLoad > 0) |>
  transmute(id, shape, scheduled = wormScheduled, realized = wormLoad,
            realization = realized / scheduled, ratio) |>
  kable(digits = 2)
```

## 10. Sensitivities

Four refits, each moving one modeling decision, compared on the daily-config
multipliers. S1 drops the practice term (if floret collapses toward its raw
geo-mean, the practice term is what separates its lattice cost from its
learning curve). S2 adds a linear whole-session drift in play order. S3 swaps
the Gaussian residual for Student-t (does a-01, or floret's wild pair, drag
anything). S4 adds a signed per-strike nuisance term (the fast-bias question
from section 3).

```{r sensitivities, results="hide"}
fit_df$seq_c <- as.numeric(scale(fit_df$seq))

refit <- function(formula, prior_obj = priors, family = gaussian()) {
  brm(formula, data = fit_df, family = family, prior = prior_obj,
      chains = 4, iter = 3000, warmup = 1500,
      control = list(adapt_delta = 0.99),
      cores = 4, seed = 20260803, refresh = 0)
}
f_s1 <- as.formula(paste("log_ratio ~",
  paste(c("0 + Intercept", core_cols, gim_cols), collapse = " + ")))
f_s2 <- update(f_main, . ~ . + seq_c)
f_s4 <- update(f_main, . ~ . + bombHits)

fit_s1 <- refit(f_s1, prior_obj = priors[priors$coef != "practice_c", ])
fit_s2 <- refit(f_s2, c(priors, set_prior("normal(0, 0.5)", class = "b", coef = "seq_c")))
fit_s3 <- refit(f_main, family = student())
fit_s4 <- refit(f_s4, c(priors, set_prior("normal(0, 0.2)", class = "b", coef = "bombHits")))
```

```{r sensitivity-table}
daily_mult_of <- function(fit, label) {
  dr <- as_draws_df(fit)
  daily_cfgs |>
    rowwise() |>
    mutate(model = label,
           mult = mean(exp(dr[[paste0("b_", SHAPE_TABLE[[shape]])]] +
                       cells * dr[[paste0("b_", SHAPE_TABLE[[shape]], "_x_cellCount")]] +
                       mines * dr[[paste0("b_", SHAPE_TABLE[[shape]], "_x_totalMines")]]))) |>
    ungroup() |> select(model, shape, mult)
}
sens <- bind_rows(
  daily_mult_of(fit_main, "main"),
  daily_mult_of(fit_s1, "S1 no practice"),
  daily_mult_of(fit_s2, "S2 + session drift"),
  daily_mult_of(fit_s3, "S3 student-t"),
  daily_mult_of(fit_s4, "S4 + strike term")
) |>
  pivot_wider(names_from = shape, values_from = mult)
kable(sens, digits = 2)

sens_stats <- map_dfr(
  list(main = fit_main, s1 = fit_s1, s2 = fit_s2, s3 = fit_s3, s4 = fit_s4),
  function(f) {
    fe <- fixef(f)
    tibble(practice = ifelse("practice_c" %in% rownames(fe),
                             fe["practice_c", "Estimate"], NA),
           drift = ifelse("seq_c" %in% rownames(fe), fe["seq_c", "Estimate"], NA),
           strike = ifelse("bombHits" %in% rownames(fe), fe["bombHits", "Estimate"], NA))
  }, .id = "model")
kable(sens_stats, digits = 3)
```

Where a shape's multiplier is stable across the five rows, the number is a
property of the data; where it moves (expect floret, and deltoidal to a
lesser degree), the moving decision is part of the answer and the prior width
should carry it. The novice end of the practice lever is not in this table
because it is not in the measured range: the fitted line covers within-shape
exposures 4 through 15, and the honest order-1 read is the warm-up table in
section 5, not an extrapolation.

## 11. What the tier ladder and the band inversion take from this

Challenge-250 prices levels in par-per-cell, and the Phase-2 shape rotation
needs per-date config tables whose fitted par lands inside the daily band (20
to 240 s, easy-skewed). The `config_tbl` above is the input to both: fitted
Greg par and par-per-cell for every config the lab actually proved
generable, at the coefficients this document fits. Three constraints ride
along:

- **Density floors are real.** Rhombille has no reachable no-guess board
  space below about 0.23 density (measured during battery design, reproduced
  at 0.18), so its equation only describes 0.23 and up; the other Laves
  lattices were proven down to 0.18 and the open lattices to 0.14. A band
  inversion must respect each lattice's proven range, and any config outside
  the lab grid needs a generation proof (the `validate-parlab-battery.mjs`
  pattern) before it ships.
- **Deltoidal's daily config prices near the band ceiling.** Its fitted daily
  par sits where the current fixed config barely fits under 240 s; the
  inversion will likely want its daily boards sparser or smaller, which the
  S0/M0 rows of the config table cover.
- **These are one practiced player's multipliers.** The tier ladder's
  absolute anchors (the numeric par-per-cell tier boundaries) should be set
  after live tiling scores exist; the map doc ships with tier indices and
  placeholder ranges on exactly this reasoning.

```{r band-check}
config_tbl |>
  mutate(in_daily_band = fitted_par >= 20 & fitted_par <= 240) |>
  group_by(shape) |>
  summarise(configs = n(), inside_band = sum(in_daily_band),
            min_fitted = min(fitted_par), max_fitted = max(fitted_par),
            .groups = "drop") |>
  kable(digits = 0)
```

## 12. The prior-centers artifact

The deviation posteriors, written as a machine-readable record beside the
data snapshot. Nothing reads this file yet; it exists so the design
conversation below has a frozen referent, pinned to the base vintage it was
measured against.

```{r write-artifact}
artifact <- list(
  format = "parlab-prior-centers-v1",
  fitDate = "2026-08-03",
  baseVintage = base_vintage,
  baseModel = PAR_MODEL[c("intercept", names(COEF_TO_PREDICTOR))],
  snapshot = SNAPSHOT_PATH,
  n_rows = nrow(fit_df),
  sigma = unname(summary(fit_main)$spec_pars["sigma", "Estimate"]),
  practiceCenter = practice_center,
  sessionPace = unname(fe["Intercept", "Estimate"]),
  practiceSlope = unname(fe["practice_c", "Estimate"]),
  deviations = post_tbl |>
    filter(term %in% dev_terms) |>
    transmute(term, mean = round(mean, 5), sd = round(sd, 5), n_rows) |>
    (\(x) setNames(pmap(list(x$mean, x$sd, x$n_rows),
                        \(m, s, n) list(mean = m, sd = s, nRows = n)), x$term))()
)
write_json(artifact, "data/parlab-prior-centers.json",
           auto_unbox = TRUE, pretty = TRUE, digits = NA)
cat("wrote data/parlab-prior-centers.json\n")
```

## 13. How these priors could enter the nightly refit

The decision this document deliberately does not make. The mechanics that
constrain every option: the nightly recomposes `PAR_MODEL_SHAPES` from its own
fit each run, so a hand-edited block does not survive the next refit; a
deviation column with no live rows is gated out of the nightly's formula
entirely (the zero-variance gate); and the shipped block zeroes any deviation
until its column has 20 nonzero fit rows (`NEW_FEATURE_DATA_THRESHOLD`).
Whatever enters must enter through `scripts/refit-par-model.R`, or the next
nightly reverts it.

**Option A: lab posteriors as the nightly's deviation prior centers.** Replace
`normal(0, INTERACTION_PRIOR_SD)` with `normal(lab_mean, width)` for the 18
core terms (and any gimmick terms wanted). The zero-variance gate is the
catch: with no live tiling rows a deviation column never enters the formula,
so its prior never matters and the composed block stays parity. Prior centers
alone change nothing until the flip produces rows, and then only 20-plus rows
at a time under the ship guard.

Sub-decision if A: the width. The lab posterior SDs assert one practiced
player generalizes; `INTERACTION_PRIOR_SD` (0.5) around the lab mean keeps
the center but lets live data walk away cheaply; something between (say
doubling the lab SD) splits the difference.

**Option B: additionally seed the shipped blocks.** Emit the lab-centered
deviations into `PAR_MODEL_SHAPES` even before live rows exist (bypassing the
zero-variance gate and the 20-row guard for lab-seeded terms, with the lab
posterior standing in for the missing likelihood). This is the only option
where the flip's first deltoidal daily ships a par near five times parity
instead of parity; the cost is shipping one player's practiced multipliers as
everyone's par until live data accumulates, and a real change to the refit's
guard logic (the guards exist to stop prior medians reaching `predictPar`,
and this deliberately routes around them for a named term set).

**Option C: lab rows as data in the nightly fit.** Append these 68 rows to
the fit frame (they are real plays under the daily time convention) with a
lab-session nuisance offset, letting the joint fit weigh them against live
rows forever. Statistically the cleanest long-run pooling; the costs are
plumbing (the nightly would fetch `parLab/` and carry embedded per-row
features, which nothing in its daily frame does today; only the separate
timed fit works that way) and permanence (his lab pace becomes a standing
fixture in the production fit rather than a prior that live data can retire).

**Option D: nothing until live data.** The pure 20-row path. The flip then
ships parity par on every Laves daily for weeks (deltoidal wins recorded at
a fifth of their real difficulty, feeding the fit distorted early rows), and
the Phase-2 band inversion has no equation to invert, which is the blocker
that motivated this battery in the first place.

A and B compose (A is the fit-side memory, B the ship-side bridge), and
either can carry the strike re-pricing iteration later. The width
sub-decision, the practiced-player caveat, and which gimmick cells deserve
anything beyond zero-centered-wide are all part of the same conversation.
