---
title: "Greg-par: refit the linear model"
author: "Christopher D. Wells"
format: html
---

This document loads a Firebase Realtime Database JSON export, joins every
daily completion to its board feature vector, fits a linear regression
predicting completion time from board features, and prints a copy-paste
`PAR_MODEL` object for the JavaScript client.

Workflow:

1. In the Firebase console, export the full database as JSON: **Realtime
   Database → ⋮ menu → Export JSON**. Save the file as `firebase-export.json`
   next to this `.qmd`.
2. Render this document (Render in RStudio, or `quarto render fit-par-model.qmd`).
3. Copy the printed `PAR_MODEL = { ... }` block into
   `src/logic/difficulty.js`. Commit the file with the refit metadata
   comment updated to the new date and sample size.
4. Deploy.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(jsonlite)
library(broom)
library(ggplot2)
library(car) # for vif()
```

## 1. Load and join

```{r load}
raw <- fromJSON("firebase-export.json", simplifyVector = FALSE)

# dailyMeta/{date}/features — one row per date
meta <- tibble(
  date = names(raw$dailyMeta %||% list()),
  features = map(raw$dailyMeta %||% list(), ~ .x$features)
) %>%
  filter(!map_lgl(features, is.null)) %>%
  unnest_wider(features)

# daily/{date}/{pushId} — one row per score
scores <- raw$daily %||% list()
scores_df <- tibble(
  date = rep(names(scores), map_int(scores, length)),
  entry = flatten(map(scores, ~ .x))
) %>%
  mutate(
    name = map_chr(entry, ~ .x$name %||% NA_character_),
    time = map_dbl(entry, ~ .x$time %||% NA_real_),
    uid  = map_chr(entry, ~ .x$uid  %||% NA_character_),
    par  = map_dbl(entry, ~ .x$par  %||% NA_real_),
    bombHits = map_dbl(entry, ~ .x$bombHits %||% 0)
  ) %>%
  select(-entry) %>%
  filter(!is.na(time), time >= 5, time <= 3600)

# Join — every score gets the board's feature vector
df <- scores_df %>%
  inner_join(meta, by = "date") %>%
  mutate(across(c(passAMoves, canonicalSubsetMoves, genericSubsetMoves,
                  advancedLogicMoves, disjunctiveMoves,
                  totalMines, cellCount, wallEdgeCount, mysteryCellCount,
                  liarCellCount, lockedCellCount, wormholePairCount,
                  mirrorPairCount, sonarCellCount, compassCellCount),
                ~ as.numeric(.)))

cat("N scores joined to features:", nrow(df), "\n")
cat("N unique dates:",  n_distinct(df$date), "\n")
cat("N unique players:", n_distinct(df$uid), "\n")
```

## 2. Fit

Linear model matching the `PAR_MODEL` coefficient names exactly.

```{r fit}
fit <- lm(
  time ~
    passAMoves + canonicalSubsetMoves + genericSubsetMoves +
    advancedLogicMoves + disjunctiveMoves +
    cellCount + totalMines + wallEdgeCount +
    mysteryCellCount + liarCellCount + lockedCellCount +
    wormholePairCount + mirrorPairCount +
    sonarCellCount + compassCellCount,
  data = df
)

summary(fit)
```

## 3. Diagnostics

```{r diagnostics}
# Residual plots
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))

# QQ plot — a pronounced heavy right tail suggests log-linear would fit
# better. If so, refit with log(time) and update predictPar in JS to
# exp(predicted) - 1 or similar.

# Multicollinearity check — VIF > ~5 on a feature means the regression is
# struggling to tease it apart from another. totalClicks and the move-type
# counts are expected to be collinear; we exclude totalClicks from the model
# for exactly this reason.
vif(fit)

# Residuals stratified by presence of hard moves — a systematic pattern
# here suggests within-bucket heterogeneity (e.g. canonical vs. generic
# subset split needs refining).
df %>%
  mutate(
    has_disjunctive = disjunctiveMoves > 0,
    has_generic    = genericSubsetMoves > 0,
    resid = residuals(fit)
  ) %>%
  group_by(has_disjunctive, has_generic) %>%
  summarise(n = n(), mean_resid = mean(resid), .groups = "drop")
```

## 4. Bootstrap coefficient CIs

```{r bootstrap}
set.seed(42)
n_boot <- 1000
coef_boot <- replicate(n_boot, {
  idx <- sample(seq_len(nrow(df)), replace = TRUE)
  coef(lm(formula(fit), data = df[idx, ]))
})

boot_cis <- apply(coef_boot, 1, quantile, probs = c(0.025, 0.975)) %>%
  t() %>%
  as_tibble(rownames = "term") %>%
  rename(lo = `2.5%`, hi = `97.5%`)

boot_cis
```

## 5. Emit JS `PAR_MODEL` block

```{r emit}
co <- coef(fit)

# Coerce negative coefficients to 0 — par should be monotonically
# non-decreasing in every feature (a mystery cell cannot make a board
# faster). If the regression says otherwise, it is probably collinearity
# or noise, not a real effect. Revisit in the diagnostics if this clamps
# fire frequently.
nn <- function(x) max(0, as.numeric(x))

js <- sprintf(
"export const PAR_MODEL = {
  // Last refit: %s | N=%d scores across %d dates, %d players
  intercept: %.2f,

  secPerPassAMove:            %.2f,
  secPerCanonicalSubsetMove:  %.2f,
  secPerGenericSubsetMove:    %.2f,
  secPerAdvancedLogicMove:    %.2f,
  secPerDisjunctiveMove:      %.2f,

  secPerCell:      %.3f,
  secPerMineFlag:  %.3f,
  secPerWallEdge:  %.3f,

  secPerMysteryCell:   %.3f,
  secPerLiarCell:      %.3f,
  secPerLockedCell:    %.3f,
  secPerWormholePair:  %.3f,
  secPerMirrorPair:    %.3f,
  secPerSonarCell:     %.3f,
  secPerCompassCell:   %.3f,
};",
  Sys.Date(), nrow(df), n_distinct(df$date), n_distinct(df$uid),
  nn(co["(Intercept)"]),
  nn(co["passAMoves"]),
  nn(co["canonicalSubsetMoves"]),
  nn(co["genericSubsetMoves"]),
  nn(co["advancedLogicMoves"]),
  nn(co["disjunctiveMoves"]),
  nn(co["cellCount"]),
  nn(co["totalMines"]),
  nn(co["wallEdgeCount"]),
  nn(co["mysteryCellCount"]),
  nn(co["liarCellCount"]),
  nn(co["lockedCellCount"]),
  nn(co["wormholePairCount"]),
  nn(co["mirrorPairCount"]),
  nn(co["sonarCellCount"]),
  nn(co["compassCellCount"])
)

cat(js, "\n")
```

Copy the printed block into `src/logic/difficulty.js`, replacing the existing
`PAR_MODEL`. Update the refit-metadata comment with today's date and
the sample size.
