---
title: "How much does the par scale move when the roster changes?"
subtitle: "Intercept sensitivity, and whether a reference-cohort anchor absorbs it"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## The question

Par is fit as `log(pure_time) ~ features + (1 | uid)`, and the intercept is
recentered so that the play-weighted geometric mean of the player handicaps is
1. That makes par the time of the *play-weighted average player in the fit*.

So the absolute level of par is a property of **who is currently playing**, not
of the boards. If three slower players join the daily roster and play
regularly, the average slows, the intercept rises, and every board's predicted
par rises with it.

That matters here because the challenge ladder is priced in **par-seconds per
cell**. Its tier anchors (T1 0.55 through T12 3.60), its 8-minute par ceiling,
and the endless zone's 3.5 s/cell admission floor are all absolute numbers in
seconds. If the second itself stretches, the ladder is being sorted and
filtered by a ruler that moved.

This document measures how far it moves, and whether pricing the ladder in
**reference-cohort seconds** — `par x k_ref`, where `k_ref` is the handicap of
a frozen set of consistent players — holds still while par does not.

Two things it deliberately does not do. It does not change the par model:
Christopher's ruling is that Greg stays the median of the complete population,
so the recentering is left exactly as it is and only the *ladder's* yardstick
is in question. And it does not use `brms`. The question is how a point
estimate of the intercept responds to cohort composition, not what the
posterior looks like; `lmer` answers that in seconds rather than minutes, and
the agreement between the two is checked below rather than assumed.

```{r setup, message=FALSE, warning=FALSE}
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(lme4)
library(ggplot2)

set.seed(20260808)

`%||%` <- function(a, b) if (is.null(a)) b else a

DB_URL <- "https://gregsweeper-66d02-default-rtdb.firebaseio.com"

# Mirrors scripts/refit-par-model.R. Kept as a literal copy rather than sourced
# because that script runs the whole nightly pipeline on load.
COEF_TO_PREDICTOR <- c(
  secPerCell         = "cellCount",
  secPerMineFlag     = "totalMines",
  secPerPatternMove  = "patternMoves",
  secPerSearchMove   = "searchMoves",
  secPerWallEdge     = "wallEdgeCount",
  secPerZeroCluster  = "zeroClusterCount",
  secPerMysteryCell  = "mysteryCellCount",
  secPerLiarCell     = "liarCellCount",
  secPerLockedCell   = "lockedCellCount",
  secPerWormholePair = "wormholePairCount",
  secPerMirrorPair   = "mirrorPairCount",
  secPerSonarCell    = "sonarCellCount",
  secPerCompassCell  = "compassCellCount",
  secPerWormLoad     = "wormLoad"
)
PREDICTORS <- unname(COEF_TO_PREDICTOR)

LEGACY_BOMB_RATE            <- 15.0
BOMB_HIT_CHEAT_FRACTION     <- 0.30
MIN_PLAYS_FOR_FIT_INCLUSION <- 5
```

## The data

Both nodes are world-readable, so this needs no credentials.

```{r fetch, cache=TRUE, message=FALSE}
meta_raw   <- fromJSON(paste0(DB_URL, "/dailyMeta.json"), simplifyVector = FALSE)
scores_raw <- fromJSON(paste0(DB_URL, "/daily.json"),     simplifyVector = FALSE)

cat(sprintf("meta rows: %d | score dates: %d\n", length(meta_raw), length(scores_raw)))
```

```{r prep}
# Features, one row per date.
meta <- imap_dfr(meta_raw, function(m, date) {
  f <- m$features
  if (is.null(f)) return(NULL)
  row <- as.list(setNames(rep(0, length(PREDICTORS)), PREDICTORS))
  # patternMoves / searchMoves are DERIVED from the stored raw move-type
  # counts, exactly as the shipped model derives them.
  raw <- function(k) if (is.null(f[[k]])) 0 else as.numeric(f[[k]])
  for (p in PREDICTORS) if (!is.null(f[[p]])) row[[p]] <- as.numeric(f[[p]])
  row$patternMoves <- raw("canonicalSubsetMoves") + raw("genericSubsetMoves")
  row$searchMoves  <- raw("advancedLogicMoves")
  row$date <- date
  as_tibble(row)
})

# Scores, one row per completion.
scores <- imap_dfr(scores_raw, function(entries, date) {
  imap_dfr(entries, function(e, id) {
    if (is.null(e$uid) || is.null(e$time)) return(NULL)
    bombBaseSum <- if (!is.null(e$bombHitEvents)) {
      sum(map_dbl(e$bombHitEvents, ~ .x$penalty %||% 0))
    } else NA_real_
    tibble(
      date     = date,
      uid      = e$uid,
      time     = as.numeric(e$time),
      bombHits = as.numeric(e$bombHits %||% 0),
      bombBase = bombBaseSum,
      hints    = length(e$hintEvents %||% list())
    )
  })
})

df <- scores |>
  inner_join(meta, by = "date") |>
  # The shipped filters: no hinted rows, no brute-force probes.
  filter(hints == 0, bombHits <= BOMB_HIT_CHEAT_FRACTION * totalMines) |>
  # Bomb cost comes OUT before logging, so there is no bomb regressor.
  mutate(
    pure_time = time - if_else(!is.na(bombBase), bombBase, bombHits * LEGACY_BOMB_RATE),
    pure_time = pmax(pure_time, 1)
  ) |>
  group_by(uid) |> filter(n() >= MIN_PLAYS_FOR_FIT_INCLUSION) |> ungroup()

cat(sprintf("usable rows: %d across %d players and %d boards\n",
            nrow(df), n_distinct(df$uid), n_distinct(df$date)))
```

## The fit, and the recentering

`fit_scale()` is the whole pipeline in miniature: fit, then recenter the random
intercepts in log space by their play-weighted geometric mean, exactly as the
nightly refit does, and report the resulting intercept.

```{r fitter}
# Only predictors that actually vary can enter; the nightly fit applies the
# same per-column zero-variance gate.
live_predictors <- PREDICTORS[map_lgl(PREDICTORS, ~ sd(df[[.x]], na.rm = TRUE) > 0)]

fit_scale <- function(data) {
  form <- as.formula(paste("log(pure_time) ~", paste(live_predictors, collapse = " + "), "+ (1 | uid)"))
  m <- lmer(form, data = data, REML = FALSE,
            control = lmerControl(calc.derivs = FALSE))

  re  <- ranef(m)$uid[, "(Intercept)"]
  ids <- rownames(ranef(m)$uid)
  # Play-weighted geometric-mean recentering, in log space.
  w      <- as.numeric(table(data$uid)[ids])
  centre <- sum(re * w) / sum(w)

  list(
    model     = m,
    intercept = unname(fixef(m)["(Intercept)"]) + centre,
    k         = setNames(exp(re - centre), ids),
    coefs     = fixef(m)[live_predictors]
  )
}

base <- fit_scale(df)
cat(sprintf("baseline intercept (log seconds): %.4f\n", base$intercept))
```

### Does `lmer` stand in for `brms` here?

The claim being leaned on is that the intercept's *response to cohort
composition* is captured by a point estimate. Check it against the shipped
model rather than asserting it: the shipped `PAR_MODEL` intercept was produced
by the real brms pipeline on this same data.

```{r brms-check}
shipped <- readLines("../src/logic/difficulty.js")
# The marker names appear in the surrounding prose too, so take the LAST match
# of each — the markers themselves.
i <- max(grep("PAR_MODEL:START", shipped))
j <- max(grep("PAR_MODEL:END", shipped))
# The first `intercept:` assignment inside the marked block. The comment
# lines around it mention the word too, so match the assignment itself.
shipped_intercept <- as.numeric(sub("^\\s*intercept:\\s*(-?[0-9.]+),.*$", "\\1",
                                    grep("^\\s*intercept:", shipped[i:j], value = TRUE)[1]))

cat(sprintf("shipped (brms) intercept: %.4f\nlmer intercept:           %.4f\ndifference:               %.4f log units (%.1f%% on par)\n",
            shipped_intercept, base$intercept,
            base$intercept - shipped_intercept,
            100 * (exp(base$intercept - shipped_intercept) - 1)))
```

A gap here is expected and is not the thing under test — the two differ by
priors, by the bias correction the pipeline applies afterwards, and by however
much data has landed since the last refit. What matters is that the *shift*
measured below is large or small relative to that gap.

## The simulation

Add `n_new` players, each `slowness`x the current average and each playing
`n_plays` boards drawn from the boards already in the data, then refit and see
where the intercept lands. Slowness is applied multiplicatively on the log
scale, which is how the model represents a player, with realistic within-player
noise so the new players are not implausibly consistent.

```{r simulate}
add_players <- function(data, n_new, slowness, n_plays, sigma = 0.25) {
  boards <- data |> distinct(date, .keep_all = TRUE) |> select(-uid, -time, -bombHits, -bombBase, -hints, -pure_time)
  # Predicted clean time for the average player on each board.
  boards$mu <- predict(base$model, newdata = boards, re.form = NA)

  map_dfr(seq_len(n_new), function(p) {
    picks <- boards[sample(nrow(boards), n_plays, replace = nrow(boards) < n_plays), ]
    picks |>
      mutate(
        uid       = paste0("sim_slow_", p),
        pure_time = exp(mu + log(slowness) + rnorm(n(), 0, sigma)),
        time      = pure_time, bombHits = 0, bombBase = 0, hints = 0
      ) |>
      select(-mu)
  }) |>
    bind_rows(data)
}

grid <- expand_grid(
  n_new    = c(1, 3, 5),
  slowness = c(1.3, 1.6, 2.0),
  n_plays  = c(20, 60)
)

# THE REFERENCE COHORT is the CONSISTENT players, not everyone already in the
# fit. Taking everyone folds in a marginal 6-play account whose handicap is
# both extreme and badly estimated, which is what made the first run of this
# document report the anchor as not working.
plays    <- sort(table(df$uid), decreasing = TRUE)
regulars <- names(plays)[plays >= 50]
cat(sprintf("reference cohort: %d players with %s plays
",
            length(regulars), paste(plays[regulars], collapse = "/")))

# The anchor's OWN baseline. Its absolute level is a one-time rescale and does
# not matter; whether it MOVES is the whole question, so every anchored figure
# is normalised by this.
k_ref_base <- exp(mean(log(base$k[names(base$k) %in% regulars])))
cat(sprintf("baseline k_ref: %.4f
", k_ref_base))

results <- pmap_dfr(grid, function(n_new, slowness, n_plays) {
  sim <- add_players(df, n_new, slowness, n_plays)
  f   <- fit_scale(sim)
  tibble(
    n_new = n_new, slowness = slowness, n_plays = n_plays,
    intercept   = f$intercept,
    par_inflate = exp(f$intercept - base$intercept),
    # The proposed anchor: the reference cohort's own predicted time.
    # k_ref is the geometric mean handicap of the players who were already
    # in the fit before the newcomers arrived.
    k_ref       = exp(mean(log(f$k[names(f$k) %in% regulars]))),
  ) |>
    mutate(anchored_inflate = par_inflate * k_ref / k_ref_base)
})

results |>
  mutate(across(c(par_inflate, k_ref, anchored_inflate), ~ round(.x, 4))) |>
  arrange(n_plays, slowness, n_new) |>
  knitr::kable(caption = "Par inflation vs the reference-cohort-anchored scale")
```

## What it means for the ladder

`par_inflate` is the factor every board's par is multiplied by, and therefore
the factor its par-per-cell is multiplied by. Three consequences, all of which
push the ladder the same way:

```{r consequences}
worst <- results |> slice_max(par_inflate, n = 1)

pool <- fromJSON("data/ladder-pool-prices.json")   # written by the block below

report <- results |>
  rowwise() |>
  mutate(
    # A board that measures exactly at the endless floor today measures this
    # afterwards; the floor admits everything down to 3.5 / inflate.
    floor_admits_from = 3.5 / par_inflate,
    # Boards that used to fit the 8-minute ceiling and no longer do.
    ceiling_lost_pct  = 100 * mean(pool$par * par_inflate > 480) - 100 * mean(pool$par > 480)
  ) |>
  ungroup()

report |>
  select(n_new, slowness, n_plays, par_inflate, floor_admits_from, ceiling_lost_pct, anchored_inflate) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = "Ladder consequences: where the endless floor really sits, and what the par ceiling drops")
```

```{r plot, fig.width=8, fig.height=4}
results |>
  pivot_longer(c(par_inflate, anchored_inflate), names_to = "scale", values_to = "factor") |>
  mutate(scale = recode(scale,
                        par_inflate = "par (population-anchored)",
                        anchored_inflate = "par x k_ref (cohort-anchored)")) |>
  ggplot(aes(factor(slowness), factor, colour = scale, shape = factor(n_new))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey40") +
  geom_point(size = 3, position = position_dodge(width = 0.4)) +
  facet_wrap(~ n_plays, labeller = label_both) +
  labs(x = "how much slower the newcomers are",
       y = "scale factor vs today",
       colour = NULL, shape = "new players") +
  theme_minimal(base_size = 12)
```

## Reading it

The dashed line is "the ruler did not move". The population-anchored series is
what the ladder is sorted by today; the cohort-anchored series is what it would
be sorted by under the proposed change. The gap between them is the whole
argument.

## What the numbers said

Measured on the live data (399 usable rows, four players at 164/159/70/6
plays, so three consistent ones):

- **Par inflates materially.** Three newcomers at 1.6x playing 60 boards each
  moves par by **x1.15**; the same three at 20 boards each move it x1.09; five
  at 2.0x playing 60 boards move it **x1.46**.
- **The endless floor is the sharpest consequence.** A 3.5 s/cell admission
  floor would, after that middle scenario, actually be admitting boards worth
  3.06 today — and 2.40 in the worst case. The zone softens by 13% to 31%
  without anything about the boards changing.
- **The par ceiling quietly deletes the large boards.** Nine percent of the
  ladder pool crosses 480s in the middle scenario, twenty percent in the worst.
- **The anchor holds.** `par x k_ref` moves between 0.93 and 1.08 across every
  scenario, mostly inside two percent, against 8-46% unanchored.

Three caveats, none of which change the direction:

1. **These are an upper bound.** The `lmer` intercept sits well above the
   shipped brms one, because the shipped figure carries priors, the pipeline's
   bias correction, per-shape deviations and other data sources. The *shift* is
   a within-estimator comparison and so is valid, but brms partial-pools
   newcomers harder than ML does, so the real pipeline should move less than
   this.
2. **The sample is tiny.** Four players, one of them with six plays. The
   baseline itself is fragile, which is part of why the ruler moves so easily.
3. **The residual is real.** The anchor is not exact — a few percent of drift
   survives it, because the newcomers change the feature coefficients slightly
   as well as the intercept, and no intercept anchor can absorb that.
