---
title: "Does par need a size term? The log(cells) question"
author: "Christopher D. Wells"
date: last-modified
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## The question

The par model prices a 187-cell sparse plain Classic at about 80 seconds,
which is the quick band, and a human who has played one knows that number is
wrong. The long honeycomb from the marathon arc made the same point from the
other side: the played time blew the model's par out of the water. The
diagnosis has been on file for a while: board size is MEDIATED through the
move counts (a bigger board has more pattern moves, and those carry the
time), so the fit leaves the direct `cellCount` coefficient near zero and the model is
blind to the one cost that scales with area rather than with decisions,
the travel and scanning time between moves. Extrapolation past the fitted
range is then absurd in both directions (a 600-cell hex priced 17 s, a
325-cell cairo priced 4.9 hours), which is why the marathon lane ships
provisional anchor-derived pars at all.

The candidate fix needs no new data: add `log(cellCount)` to the log-scale
formula, so par carries an elasticity in board size. On the log scale the
current model is

$$\log T = \alpha + \beta_c \cdot \text{cells} + \sum_j \beta_j x_j$$

and the proposal is

$$\log T = \alpha + \gamma \log(\text{cells}) + \beta_c \cdot \text{cells} + \sum_j \beta_j x_j$$

with $\gamma$ free. $\gamma = 0$ reproduces the current form; $\gamma = 1$
means par scales proportionally with cells at fixed per-cell difficulty. The
decision this document feeds: whether $\gamma$ earns a place in `PAR_MODEL`,
which re-prices every stored artifact the night it ships.

```{r setup}
# The frame is SHARED with par-model-move-rates.qmd rather than built inline
# here. It used to live in this document; it was extracted on 2026-08-20 when
# a second decision needed the same frame, because two documents mirroring the
# pipeline separately drift apart the first time the pipeline moves, and a
# stale mirror still reads like evidence. What it promises, and the crosscheck
# that keeps it honest against modelHistory's own n_scores, are in
# scripts/par-model-frame.R.
source("par-model-frame.R")
library(brms)
library(loo)
```

```{r pull}
#| cache: true
raw <- pull_par_data()
frame <- build_par_frame(raw)
df_fit <- frame$df_fit
c(fit_rows = nrow(df_fit), eligible = length(frame$eligible))
```

## The two fits

Both are the pipeline's own flat path: `log(pure_time)` on the base
predictors plus the shape indicators, a `(1|uid)` random intercept, and
weakly informative priors. The shipped fit's non-linear split and its lb = 0
blanket exist to bound board-feature coefficients; the size question is
isolated here on the unbounded path deliberately, because $\gamma$'s sign is
the question and a bound would beg it. The shape indicators enter as plain
columns so lattice cell counts cannot masquerade as a size effect: lattices
are the small-cell-count boards, and without shape terms $\gamma$ would soak
their intercept differences.

```{r fits}
#| cache: true
sh <- add_shape_columns(df_fit)
df_fit2 <- sh$df
shape_names <- sh$names

base_terms <- c(
  "cellCount", "totalMines", "patternMoves", "searchMoves",
  "wallEdgeCount", "mysteryCellCount", "liarCellCount", "lockedCellCount",
  "wormholePairCount", "mirrorPairCount", "sonarCellCount",
  "compassCellCount", "zeroClusterCount"
)
if (any(df_fit2$wormLoad > 0)) base_terms <- c(base_terms, "wormLoad")
if (length(unique(df_fit2$archivePlay)) > 1) base_terms <- c(base_terms, "archivePlay")
if (length(unique(df_fit2$matchPlay)) > 1) base_terms <- c(base_terms, "matchPlay")
base_terms <- c(base_terms, shape_names)

f0 <- as.formula(paste("log(pure_time) ~", paste(base_terms, collapse = " + "), "+ (1|uid)"))
f1 <- update(f0, . ~ . + logCells)

m0 <- brm(f0, data = df_fit2, family = gaussian(),
          chains = 4, iter = 2000, cores = 4, seed = 20260818,
          control = list(adapt_delta = 0.95), refresh = 0)
m1 <- brm(f1, data = df_fit2, family = gaussian(),
          chains = 4, iter = 2000, cores = 4, seed = 20260818,
          control = list(adapt_delta = 0.95), refresh = 0)
```

```{r gamma}
# The one estimate this document exists to report.
post <- as_draws_df(m1)
gamma <- post$b_logCells
cat(sprintf("gamma = %.3f [%.3f, %.3f], P(gamma > 0) = %.3f\n",
            median(gamma), quantile(gamma, .025), quantile(gamma, .975),
            mean(gamma > 0)))
```

```{r shifts}
# What the size term takes from the others: the mediation prediction is that
# cellCount (already near-inert) stays near zero, the intercept drops toward
# something interpretable, and the move-count coefficients give up the part
# of their weight that was standing in for traversal.
fe0 <- fixef(m0)[, "Estimate"]
fe1 <- fixef(m1)[, "Estimate"]
shared <- intersect(names(fe0), names(fe1))
tibble(term = shared, m0 = fe0[shared], m1 = fe1[shared],
       shift = fe1[shared] - fe0[shared]) |>
  arrange(desc(abs(shift))) |> print(n = 20)
```

```{r residuals-by-size}
# The felt defect, as a picture: current-model residuals against cell count.
# If the model is blind to size, the residuals trend positive with cells;
# M1 should flatten them.
df_fit2 |>
  mutate(r0 = residuals(m0)[, "Estimate"],
         r1 = residuals(m1)[, "Estimate"],
         bin = cut(cellCount, breaks = c(0, 60, 90, 120, 160, 200, Inf))) |>
  group_by(bin) |>
  summarize(n = n(), mean_resid_m0 = mean(r0), mean_resid_m1 = mean(r1)) |>
  print()
```

```{r extrapolation}
# The sanity table: what each form prices at the sizes that matter. The
# sparse plain Classic at the fit ceiling (187 cells), at the marathon
# envelope (660), and the two recorded absurdities. Predictions are
# population-level (no uid), medians on the seconds scale.
cases <- tibble(
  label = c("17x11 sparse plain Classic (fit ceiling)",
            "30x22 sparse plain Classic (marathon envelope)",
            "hex ~600 cells (the 17s absurdity)",
            "cairo ~325 cells (the 4.9h absurdity)"),
  cellCount = c(187, 660, 600, 325),
  totalMines = c(26, 92, 84, 45),
  patternMoves = c(30, 105, 95, 52),  # scaled from observed per-cell rates
  searchMoves = c(2, 7, 6, 4),
  tiling = c("rect", "rect", "hex", "cairo")
)
newd <- cases |>
  mutate(logCells = log(cellCount),
         wallEdgeCount = 0, mysteryCellCount = 0, liarCellCount = 0,
         lockedCellCount = 0, wormholePairCount = 0, mirrorPairCount = 0,
         sonarCellCount = 0, compassCellCount = 0, zeroClusterCount = 3,
         wormLoad = 0, archivePlay = 0, matchPlay = 0)
for (sc in shape_names) newd[[sc]] <- as.integer(paste0("shape_", newd$tiling) == sc)
p0 <- exp(fitted(m0, newdata = newd, re_formula = NA)[, "Estimate"])
p1 <- exp(fitted(m1, newdata = newd, re_formula = NA)[, "Estimate"])
tibble(case = cases$label, current = round(p0), with_size_term = round(p1)) |> print()
# CHECK against reality: the marathon lane's anchor pricing puts the 660-cell
# sparse Classic at roughly 264-283s, and his played long honeycomb ran far
# past its par. The better form is whichever column reads nearer those facts.
```

```{r loo}
#| cache: true
loo0 <- loo(m0)
loo1 <- loo(m1)
loo_compare(loo0, loo1)
```

## The replace-form alternative (M2)

M1 ADDS the log term beside the linear one; the natural objection is that
the two are near-collinear (r = 0.95 over the fit range) and one term should
do. M2 tests that reading by REPLACING `cellCount` with `logCells`. If the
concavity were an artifact of the ADD-form's collinearity, M2 would find the
same gamma with a cleaner posterior; instead it found almost nothing
(gamma 0.15 [-0.01, 0.31]), because forced to carry the whole size axis
alone, the log term has to compromise between the small-board curvature and
the large-board level, and lands where it explains neither. Its
extrapolation stays absurd (the marathon envelope prices ~4.8 days) and its
LOO trails M1 by 8.5 (se 4.7). The pair is genuine curvature, not a
collinearity accident, which is why BOTH terms ship.

```{r m2-replace}
#| cache: true
f2 <- update(f0, . ~ . - cellCount + logCells)
m2 <- brm(f2, data = df_fit2, family = gaussian(),
          chains = 4, iter = 2000, cores = 4, seed = 20260818,
          control = list(adapt_delta = 0.97), refresh = 0)
post2 <- as_draws_df(m2)
cat(sprintf("M2 (replace): gamma = %.3f [%.3f, %.3f]\n",
            median(post2$b_logCells), quantile(post2$b_logCells, .025),
            quantile(post2$b_logCells, .975)))
cat("M2 intercept:", round(fixef(m2)["Intercept", "Estimate"], 3), "\n")
p2 <- exp(fitted(m2, newdata = newd, re_formula = NA)[, "Estimate"])
print(tibble(case = cases$label, m0 = round(p0), m1_add = round(p1),
             m2_replace = round(p2)))
loo2 <- loo(m2)
print(loo_compare(loo0, loo1, loo2))
df_fit2 |>
  mutate(r2 = residuals(m2)[, "Estimate"],
         bin = cut(cellCount, breaks = c(0, 60, 90, 120, 160, 200, Inf))) |>
  group_by(bin) |> summarize(n = n(), mean_resid_m2 = mean(r2)) |> print()
```

## What M1 does to the handicaps

His question after the M1 call: "How does that affect people's handicaps?"
A handicap is the play-weighted-recentered random intercept, so the global
intercept collapse cancels and only the RE-PARTITION across players moves k.
Measured: the established players move within a percent (the-anemone-guy
-0.2%, Kate -0.1%, Sebas +0.6%, MJP -1.3%); Hungry Time, whose 11 rows lean
on mid-size boards the old form mispriced, moves +4.5% (k 1.38 to 1.44,
rating -22.5 to -26.2). Nobody's rating changes sign or tier.

```{r handicap-shift}
#| cache: true
names_raw <- fromJSON(paste0(DB, "/playerNames.json"), simplifyVector = FALSE)
name_of <- function(u) {
  n <- names_raw[[u]]$name
  if (is.null(n)) paste0(substr(u, 1, 6), "...") else n
}
k_table <- function(fit, df) {
  re <- ranef(fit)$uid[, "Estimate", "Intercept"]
  w <- df |> count(uid) |> filter(uid %in% names(re))
  w <- setNames(w$n, w$uid)[names(re)]
  centered <- re - weighted.mean(re, w)
  pmin(pmax(exp(centered), 0.1), 10)
}
k0 <- k_table(m0, df_fit2)
k1 <- k_table(m1, df_fit2)
tibble(
  player = map_chr(names(k0), name_of),
  rows = as.integer(table(df_fit2$uid)[names(k0)]),
  k_current = round(k0, 3),
  k_m1 = round(k1[names(k0)], 3),
  k_shift_pct = round(100 * (k1[names(k0)] / k0 - 1), 1),
  rating_current = round((1 - k0) * 60, 1),
  rating_m1 = round((1 - k1[names(k0)]) * 60, 1)
) |> arrange(desc(rows)) |> print()
```

```{r collinearity}
# The mediation in one number: log(cells) against the move counts. High
# correlation is EXPECTED (it is the diagnosis); what matters is whether the
# posterior still separates them, which the gamma interval above answers.
df_fit2 |>
  summarize(cor_pattern = cor(logCells, patternMoves),
            cor_search = cor(logCells, searchMoves),
            cor_cells = cor(logCells, cellCount)) |> print()
```

## Decision sheet

His call, 2026-08-18: "Sounds like the M1 is the way forward." What that
decision rests on, from the runs above:

- $\gamma$ = 0.93 [0.54, 1.33], separated from the move counts despite the
  expected collinearity; the intercept dissolves (3.02 to -0.18, 20.5s to
  0.84s), which is the mediation diagnosis confirmed.
- The extrapolation table: M1 prices the marathon envelope at ~2 hours where
  the current form says 5.5 days, and the residual bins flatten. M1's
  extrapolation is still NOT licensed for pricing (the mines term alone
  explodes at envelope scale), so the marathon lane's anchor pricing stays.
- M2 (replace) is refuted above; the handicap re-partition is within a
  percent for established players.

What shipped (branch `cwells/par-size-term`): `secPerLogCell = "logCells"`
in `COEF_TO_PREDICTOR`, the frame and `apply_par_model` derive `logCells`
from the stored `cellCount` (never a stored feature, no `FEATURES_EPOCH`
move), and BOTH size terms ride the signed dev nlpar (`SIZE_DEV_COLS`,
prior normal(0, 1)) because M1's linear half is negative and the bounded
base block would censor it. The JS side derives the same predictor in
`COEF_TERMS` and ships the key at 0 in every `difficulty.js` block, so
`predictPar` is byte-identical until the first nightly emits the fitted
value and re-prices the libraries under the new form.
