---
title: "Should move counts enter par as rates per cell?"
author: "Christopher D. Wells"
date: last-modified
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## The question

M1 put `log(cellCount)` into the model on 2026-08-18 and the size question
looked settled. The first marathon rows, played two days later, say it is not.
The shipped model prices a 660-cell board that two people finished in about
twenty minutes at somewhere between five and thirty-one hours, depending on
which feature subset does the pricing.

M1 is not what fails there. Its two size terms very nearly cancel at that
scale: $0.7753 \times \log(660) = 5.04$ against $-0.00741 \times 660 = -4.89$.
Decomposing the prediction names the real source. On that board the move terms
contribute 61,707 seconds of pattern and 28,753 of search; on an ordinary
144-cell board the same two contribute 111 and 39, and the total lands within
4% of what people actually ran.

The mechanism is arithmetic rather than subtle. Move counts are *counts*, they
scale roughly with board area, and on the log scale each one multiplies. A
board four times larger carries about four times the moves, so par grows like
$e^{4\beta n}$ rather than like $4 \times$ anything. The candidate fix is to
divide the counts that scale with the board by the board:

$$\log T = \alpha + \gamma \log(\text{cells}) + \beta_p \frac{\text{pattern}}{\text{cells}} + \beta_s \frac{\text{search}}{\text{cells}} + \beta_m \frac{\text{mines}}{\text{cells}} + \sum_j \beta_j x_j$$

Size is then carried once, by $\gamma$, and the move terms describe *per-cell
difficulty*, which is a quantity that has no reason to grow with the board.

A preliminary OLS over the same rows put the rate form ahead on both counts:
residual SD 0.4745 against 0.5403, and marathon predictions on the right scale
rather than 5 to 25 times off in both directions. That comparison used player
fixed effects, a reduced feature set, and no partial pooling. **This document
asks whether the margin survives the frame that would actually ship.** It
changes nothing; it produces a number to rule on.

```{r setup}
# The frame is SHARED with par-model-size-offset.qmd (the M1 decision) rather
# than copied, because two documents that mirror the pipeline separately drift
# apart the first time the pipeline moves, and a stale mirror still reads like
# evidence. See scripts/par-model-frame.R for what it promises.
source("par-model-frame.R")
library(brms)
library(loo)
```

```{r pull}
#| cache: true
raw <- pull_par_data()
frame <- build_par_frame(raw)
df_fit <- frame$df_fit
c(fit_rows = nrow(df_fit), players = length(frame$eligible),
  smallest = min(df_fit$cellCount), largest = max(df_fit$cellCount))
```

## First, a defect the screen is causing

Before fitting anything: the pre-fit outlier screen rejects any row whose time
falls below `max(5, 0.3 x predicted par)`. It exists to catch probe runs. On a
board the model cannot price, the floor it computes is nonsense, and it throws
away exactly the rows that would fix the pricing.

```{r screen-damage}
frame$screened_out |>
  transmute(shape = tilingType, cells = cellCount, mines = totalMines,
            played = round(pure_time), predicted = round(predicted),
            screen_floor = round(screenFloor)) |>
  arrange(desc(cells)) |> print(n = 30)
cat("\nmarathon rows surviving the screen:",
    sum(df_fit$cellCount > 400), "of",
    sum(df_fit$cellCount > 400) + sum(frame$screened_out$cellCount > 400), "\n")
```

This is the anti-cheat lesson from 2026-08-09 repeating in a different gate: a
filter that drops data censors its own calibration set. The screen is not
wrong to exist, and the fix is not to widen it. The fix is to price boards
correctly, which is what the rest of this document is about. Worth noting
separately: honest fast plays on small lattices are refused too, where the
model over-prices in the other direction.

## The three forms

All three use the pipeline's flat path: `log(pure_time)` regressed on board
predictors, shape indicators, a `(1|uid)` random intercept, and weakly
informative priors. The shipped fit's non-linear split and its `lb = 0`
blanket bound board-feature coefficients; the question here is one of
functional form, and a bound on a rate would beg it, so the comparison runs
unbounded exactly as the M1 document did.

The gimmick and structure predictors stay COUNTS in every form. They are
placed by the generator, not derived from board area, and a board with four
compass cells really is different from one with one.

```{r fits}
#| cache: true
sh <- add_shape_columns(df_fit)
d <- sh$df
shared <- shared_terms(d, sh$names)

# M1: what ships today. Raw counts, linear cells, log cells.
f_m1 <- as.formula(paste(
  "log(pure_time) ~ cellCount + logCells + totalMines + patternMoves + searchMoves +",
  paste(shared, collapse = " + "), "+ (1|uid)"))

# R1: the counts that scale with the board become rates; size is carried once.
f_r1 <- as.formula(paste(
  "log(pure_time) ~ logCells + mineRate + patternRate + searchRate +",
  paste(shared, collapse = " + "), "+ (1|uid)"))

# R2: R1 plus the linear cell term back, so the data can say whether size
# still wants a second, non-logarithmic channel once the rates are in.
f_r2 <- update(f_r1, . ~ . + cellCount)

FIT <- function(f, seed = 20260820) brm(
  f, data = d, family = gaussian(), chains = 4, iter = 2000, cores = 4,
  seed = seed, control = list(adapt_delta = 0.95), refresh = 0)

m_m1 <- FIT(f_m1); m_r1 <- FIT(f_r1); m_r2 <- FIT(f_r2)
```

```{r coefficients}
# The rate coefficients, which are the quantities the form exists to estimate.
fixef(m_r1)[c("logCells", "mineRate", "patternRate", "searchRate"), ] |> round(4)
```

```{r loo}
#| cache: true
loo_m1 <- loo(m_m1); loo_r1 <- loo(m_r1); loo_r2 <- loo(m_r2)
loo_compare(loo_m1, loo_r1, loo_r2)
```

```{r sigma}
# Residual spread on the log scale, and what it means as a typical error.
tibble(
  form = c("M1 (shipped: raw counts)", "R1 (rates)", "R2 (rates + linear cells)"),
  sigma = c(summary(m_m1)$spec_pars["sigma", "Estimate"],
            summary(m_r1)$spec_pars["sigma", "Estimate"],
            summary(m_r2)$spec_pars["sigma", "Estimate"])
) |> mutate(typical_error = sprintf("%.0f%%", 100 * (exp(sigma) - 1))) |> print()
```

## The test that decides it: predict a board you have never seen

In-sample fit and LOO both reward a form for describing the boards in the
data. Neither asks the question the marathon lane actually poses, which is
whether a form can price a board four times larger than anything it was fitted
on. So: refit each form on boards of 250 cells or fewer, then predict every
row on a board past 400 cells, INCLUDING the six the screen refused.

```{r holdout}
#| cache: true
train <- d |> filter(cellCount <= 250)
held <- bind_rows(
  df_fit |> filter(cellCount > 400),
  frame$screened_out |> filter(cellCount > 400)
)
# The held-out rows need the same shape columns the training frame carries.
for (sc in sh$names) held[[sc]] <- as.integer(paste0("shape_", held$tilingType) == sc)

cat("train:", nrow(train), "rows up to", max(train$cellCount), "cells | held out:",
    nrow(held), "rows from", min(held$cellCount), "to", max(held$cellCount), "cells\n")

# Refit on the training subset. `update` with newdata re-runs the sampler, so
# these three genuinely never see a board past 250 cells; none is a re-read of
# a fit that already did.
h_m1 <- update(m_m1, newdata = train, refresh = 0)
h_r1 <- update(m_r1, newdata = train, refresh = 0)
h_r2 <- update(m_r2, newdata = train, refresh = 0)
```

```{r holdout-scores}
pred_pop <- function(m, nd) exp(fitted(m, newdata = nd, re_formula = NA,
                                      allow_new_levels = TRUE)[, "Estimate"])
res <- held |>
  transmute(cells = cellCount, mines = totalMines, played = round(pure_time),
            # What the lane actually showed the player: anchor pricing, not
            # predictPar. It is the incumbent, and a candidate form that cannot
            # beat it has not earned the right to replace it out here.
            lane = round(shown_par),
            M1 = round(pred_pop(h_m1, held)),
            R1 = round(pred_pop(h_r1, held)),
            R2 = round(pred_pop(h_r2, held))) |>
  arrange(desc(cells), desc(mines))
print(res, n = 20)

# One number per form: the median absolute error in LOG space, which treats
# being 10x too slow and 10x too fast as the same size of mistake. On a
# quantity that spans four orders of magnitude, that is the honest metric.
res |>
  summarize(across(c(lane, M1, R1, R2),
                   ~ sprintf("%.2f  (typical miss %.1fx)",
                             median(abs(log(.x / played))),
                             exp(median(abs(log(.x / played))))))) |>
  print()
```

## What it would cost

```{r handicap-shift}
# Handicaps are exp(play-weighted recentered random intercept), so a change of
# form re-partitions every player's rating. M1's re-partition moved
# established players by under 1% and one low-volume player by 4.5%; this is
# the same table for the rate form. A large shift is not a veto, but it is a
# thing to know before shipping rather than after.
k_of <- function(m) {
  r <- ranef(m)$uid[, "Estimate", "Intercept"]
  w <- df_fit |> count(uid) |> arrange(match(uid, names(r))) |> pull(n)
  exp(r - sum(r * w) / sum(w))
}
tibble(uid = substr(names(k_of(m_m1)), 1, 8),
       plays = df_fit |> count(uid) |> arrange(match(uid, names(k_of(m_m1)))) |> pull(n),
       k_now = round(k_of(m_m1), 3), k_rates = round(k_of(m_r1), 3)) |>
  mutate(shift = sprintf("%+.1f%%", 100 * (k_rates / k_now - 1))) |> print()
```

```{r residuals-by-size}
# Does either form still trend with size? A form blind to a real size cost
# leaves residuals sloping across the cell-count bins.
d |>
  mutate(r_m1 = residuals(m_m1)[, "Estimate"],
         r_r1 = residuals(m_r1)[, "Estimate"],
         bin = cut(cellCount, breaks = c(0, 60, 90, 120, 160, 200, Inf))) |>
  group_by(bin) |>
  summarize(n = n(), mean_resid_M1 = round(mean(r_m1), 3),
            mean_resid_R1 = round(mean(r_r1), 3)) |> print()
```

## Reading the result

The decision rests on four things, and they should be read together rather
than one at a time:

1. **LOO.** If R1 does not at least match M1 on boards already in the data,
   the form is buying extrapolation with accuracy where people actually play,
   and that is a bad trade.
2. **The holdout.** This is the one the marathon lane cares about. A typical
   miss under about 1.5x would make stored anchor pricing a fallback rather
   than the only honest option.
3. **The handicap shift.** Ratings are visible to players. A re-partition on
   the scale M1 produced is unremarkable; a large one needs its own decision.
4. **The residual trend.** Extrapolation can come right while the residuals
   still slope across the size bins, which would mean the size cost is still
   unabsorbed.

Whatever this says, the outlier screen needs attention on its own account: it
is currently refusing six real rows on the strength of a prediction nobody
believes, and no change of functional form makes a screen safe that trusts an
extrapolation without checking it.

One thing this document deliberately cannot settle. The traversal cost, the
time spent moving around a board rather than thinking about it, is the reason
a big board is slow in a way per-cell difficulty does not capture, and no row
in this frame carries it yet: `scrolled` is stated on zero of them, since it
began being written on 2026-08-18 and every earlier row is honestly absent.
The rate form is a better shape for the data that exists. It is not a
substitute for measuring the thing itself.
